Network Data Model and BerlinMOD Benchmark

### 2010-07-22 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
runtimes and paper updated

\documentclass[a4paper]{article}
%\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{float}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{todonotes}
\usepackage{rotating}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data
models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two
representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\listoftodos
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data
models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis
,HERMESMDCPelekis}
for DMFS and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,
NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

Why do we spend time on NCDM, if everything can be represented by DMFS?
Now, it is natural to give positions related to the street
network instead of x,y-coordinates. NCDM are expected to use less storage space,
because geographical information's about street curves are stored only once in
the network, whereas in DMFS each street curve is stored in each moving point
object using this street. NCDM can support query processing with specialized indexes
using their knowledge of the underlying network. It is much easier to
formulate queries about the relationships between moving objects and the network
in the NCDM. And not at last, the results of our experiments show that
our network constrained data model outperforms our data model of free movement in
two dimensional space by orders of magnitude. The network constrained data model
uses less than 60\% of the storage space and less than 50\% of the total query
run time of the data model of free movement in space, which we used in our
experiments.
We think that these results show that it is useful to develop specialized data
models for specialized data structures like NCDM for
network constrained moving objects to save storage space and reduce query run times.

@@

\cite{MOLRaQSu} proposes an incomplete abstract DMFS. Basic idea is that
spatio-temporal data types can be modeled by linear constraints and queries are
formulated using formulas from differential geometry.

\cite{RepresentingMovingObjectsGueting} proposes an abstract DMFS including the
idea of an time sliced representation for moving objects. This basic idea of
time sliced representation is used by the DMFS \cite{STAUPelekis} and
\cite{DataModelDataStructureGueting}. The last one is used in our experiments
and therefore reviewed in more detail in Section \ref{sec:bmodbdatamod}.
The main difference between the both data models is that
\cite{DataModelDataStructureGueting} supports only linear interpolation of movement,
whereas \cite{STAUPelekis} also supports arc interpolation of movement.
\cite{STAUPelekis} uses only one spatial object
containing all spatial geometries for the representation of spatial objects and
one moving object for the representation of the different moving object data types,
whereas \cite{DataModelDataStructureGueting} uses different spatial and moving
objects for the representation of the different spatial and moving data types.
According to this \cite{STAUPelekis} provides only a single operator that
distinguishes
between the different topological relationships via a parameter, whereas
\cite{DataModelDataStructureGueting} uses different operations to estimate
topological relationships. Overall, \cite{STAUPelekis} offers a more flexible
object oriented design than \cite{DataModelDataStructureGueting}.

@@

The most NCDM use edge based graph representations for the representation of the
underlying network data only a small number of NCDM uses route oriented data models
or combine route and edge based data models.

\cite{NetworkVazirgiannis} proposes an edge based NCDM. The edges and their
attributes
are stored in a relation representing the network as an undirected graph. Moving
objects are assumed to drive always on the path with the lowest cost,
in terms of distance or travel-time. They are defined by the source point, the
target point, and the starting time instant of the trip. The trajectory is
computed by this assumption using the length and speed attributes of the graph
edges within the shortest respectively fastest path computation.
The advantage of this definition is the reduced storage space for the representation
of moving point objects. The drawback is the high computational effort for query
evaluation on moving objects.

@@

Another two-layer network representation is proposed by
\cite{DynamicTransportNetworkDing}.
The authors of \cite{DynamicTransportNetworkDing} combine the advantages of the
dynamic edge-based \cite{DynNetworkModEdgeGueting} and the dynamic route-based
\cite{DynNetworkModRouteGueting} NCDM approaches. The route-based environment
reduces the update intervals and used storage space for the database representation
of moving point objects, whereas the edge-based environment supports
a more detailed view on the traffic conditions of the different edges belonging
to the same route. Moving objects are represented by a set of pairs. The pairs
consist of an motion vector and an Boolean flag. The Boolean flag tells if the
motion vector contains current or historical information. Each motion vector
consists, in parts similar to \cite{NetworkJensen}, of a time stamp, a network
position, and a speed vector. Similar to \cite{NetworkGueting} the network
position bases on routes and junctions and not on edges. Different from all
other NCDM in this section \cite{DynamicTransportNetworkDing}
uses time depending dynamic attributes in the representation of the network parts.
Therefore, changes in the network environment can be handled without loss of
information in this NCDM.

To the best of our knowledge only a few of the proposed data models have been
implemented into database management systems: \cite{STAUPelekis} is implemented
as data cartridge \cite{HERMESMDCPelekis, HERMESPelekis} for the commercial
Oracle\textregistered object-relational DBMS \cite{oracle};
\cite{DynamicTransportNetworkDing} is implemented as extension of the open source
database project PostgreSQL \cite{PostgreSQL}; \cite{DataModelDataStructureGueting}
and \cite{NetworkGueting} are implemented in the freely available extensible
\secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}.

Although \cite{STAUPelekis} for the DMFS and \cite{DynamicTransportNetworkDing} for
the NCDM provide greater flexibility we decided to use
\cite{DataModelDataStructureGueting} and \cite{NetworkGueting} in our
experiments, because both data models are available in the same DBMS, which is
also the DBMS in which the \bmodb{} \cite{BerlinMODVLDBDuentgen} has bee developed.

@@

Therefore, several data generators have been developed. Some of them generate
only unconstrained moving point objects like \cite{OportoSaglio}, or support
only short-term observations like \cite{BrinkhoffsDataGenerator}, or are
unavailable respectively require additional software like \cite{STACTS}.

In the sequel we give short reviews of the \secondo{} DBMS
(Section \ref{sec:secondo}), the both data models we used in our experiments
(Section \ref{sec:bmodbdatamod} and Section \ref{sec:netdatamod}), and the \bmodb{}
(Section \ref{sec:bmodb}).
\subsection{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use these plugins to extend
their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including SPACE (Section \ref{sec:bmodbdatamod}) and NET
(Section \ref{sec:netdatamod}). Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version 3.0.
\subsection{Data Model of \bmodb{} (SPACE)}
\label{sec:bmodbdatamod}
\cite{STDTModelingQueryingMOinDErwig} presents the basic idea of the DMFS that
is used by the \bmodb{}. The abstract data model for SPACE was pubished in
\cite{RepresentingMovingObjectsGueting} and the discrete data model in
\cite{DataModelDataStructureGueting}. Abstract data models are useful as conceptual
models, but they cannot be implemented, because computers can only use finite sets.
In the sequel description we will mainly focus on the discrete data model.

@@

The carrier sets for all data types in the discrete data model contain $\perp$,
representing an undefined value. This is not mentioned any more in the sequel.

The carrier sets for the BASE data types in the discrete data model are defined
by the corresponding programming language data types \dt{int}, \dt{real}, \dt{bool},
and \dt{string}. The carrier set of a value of the data type \dt{point} is
\dt{real} $\times$ \dt{real}. The two \dt{real} values represent the x,y-coordinates
of the \dt{point} value in the two dimensional plane. The data type \dt{points}
consists of a disjoint set of \dt{point} values. The carrier set for the data
type \dt{line}
consists of a finite set of disjoint line segments representing the linear
approximation of the line curve in the two dimensional plane. Semanticaly an
\dt{line} value is the union of the points of all its line segments. A region
is the union of all points covered by the region. The carrier set \dt{region}
is defined to be a finite set of line segments building a polygon representing
the linear approximation of the outer and, if the region contains wholes, inner
borders of the region. The borders are defined to belong to the region.

@@

The type constructor \dt{range} converts BASE and TIME data types $\alpha$ into
a type whoes values are finite sets of intervals over $\alpha$. Range types are
used to represent collections of time intervals, or the values taken by a moving
real. Intervals are represented by their start and end point and two flags
indicating
if the start respectively end point is part of the interval or not.

The other important type constructor of the abstract data model is \dt{moving}.
In the abstract data model\dt{moving} maps each data type $\alpha$ from BASE and
SPATIAL into an time dependend spatio-temporal moving data type
\dt{moving}($\alpha$)
(\dt{m$\alpha$} for short) of kind TEMPORAL. The discrete data model introduces some
additional type constructors to implement the moving type constructor of the
abstract data model. A detailed description of the type constructors is skiped
due to place limitations. We explain the realisation of \dt{moving} at the example
of a \dt{moving}(\dt{point}) (short \dt{mpoint}) object. An \dt{mpoint} value
may represent a car, which changes its position in the plain within time.

An \dt{mpoint} consists of a set of so called \dt{unit}(\dt{point}) values
(\dt{upoint} for short).
Each \dt{upoint} consists of a time interval and two \dt{point} values. The first
\dt{point} value represents the position of the \dt{upoint} at the start of the
time interval and the second \dt{point} value represents the position of the
\dt{upoint} at the end of the time interval. It is assumed that the object
represented
by the \dt{upoint} moves on the straight line between these two points with constant
speed within the given time interval. The velocity of the object is given by the
ratio from the distance of the two points and the length of the time interval of
the \dt{upoint}.

@@

This spatio-temporal data model of \dt{moving} allows us to compute the position
of an \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assuming the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by an \dt{intime(point)}
(short form \dt{ipoint}). An \dt{ipoint} consists of an time instant and an
\dt{point}
value and represents the position of the \dt{mpoint} value at the given time
instant.

Other data types of \secondo{} which are used in the \bmodb{} are \dt{mbool},
\dt{mreal}, and \dt{periods}. A \dt{mbool} value consists of a set of \dt{ubool}
values. Each \dt{ubool} value is is constant \true{} or \false{} for the given
time interval. A \dt{mreal} value consists of a set of \dt{ureal} values. Each
\dt{ureal} value is defined by a function of time representing the \dt{real}
value at each time instant. A \dt{periods} value is a set of disjoint an not
connected time intervals.
\subsection{Network Data Model(NET)}
\label{sec:netdatamod}
The central idea of the NDM presented in \cite{NetworkGueting} is that
every movement is constrained by a given network and every position can be described
relative to this network. Contrary to the most other NCDM NET models the network
in terms of routes, corresponding to roads or highways in real life. Positions
are given by a route identifier and the distance from the start of the route.
This is a more natural representation of network positions as the directed graph
representation of networks, where junctions are vertexes and the pieces between
junctions are represented by edges, which is used in the most NCDM. We have names
for roads not for junctions or pieces between junctions.

The routes based network representation has also the advantage that the
representation of moving objects that move over several sections of the same route
with constant speed becomes much smaller, because we only have to store new
information if the moving object changes the road or the speed. Not every time
it passes a junction like in the other NCDM.

In NET the data type \dt{network} is modeled by two main components. One is the
set  of routes (streets) and the other one the set of junctions (crossings). The
domain of routes is defined as
\[
\begin{split}
Route = \{(id,\ l,\ c,\ kind,\ start)\ &|\ id \in \dt{int},\ l \in \dt{real},
\ c \in \dt{line},\\
&\ kind \in \{simple,\ dual\},\\
&\ start \in \{smaller,\ larger\}\},
\end{split}
\]
where $id$ is a distinct route idenitfier, $l$ is the length of the route, $c$
is the route curve as \dt{line} value (see Section \ref{sec:bmodbdatamod}),
$kind$ indicates if the lanes of the route are separated, and $start$ indicates
how the route curve is embedded into space.

If $R$ is a set of distinct routes, the domain of junctions in $R$ is defined as
\[
  \begin{split}
    Junction(R)=\{(rm_1,\ rm_2,\ cc)\ &|\ rm_1,\ rm_2 \in RMeas(R),\\
&rm_1 = (r_1,\ d_1),\ rm_2 = (r_2, d_2),\\
&r_1 \neq r_2, cc \in \dt{int}\}.
  \end{split}
\]
Where the set of possible positions in $R\ RMeas(R)$ is defined as
\[
  \begin{split}
    RMeas(R) = \{(rid, d)\ &|\ rid \in \dt{int},\ d \in \dt{real},\\
&\exists (rid,\ l,\ c,\ k,\ s) \in R \wedge 0 \leq d \leq l\},
  \end{split}
\]
and $d$ is the distance from the start of the route.
The connectivity code $cc$ encodes which lanes of the roads are connected
by the junction\footnote{See \cite{NetworkGueting} for detailed explanations.}.

A network $N$ is a pair $(R,\ J)$, where $R$ is a finite set of distinct routes
and $J$ is a finite set of junctions in $R$. The carrier set for network
positions $Loc(N)$ is equal to the set of route locations and defined as
\[
  \begin{split}
    RLoc(R)=\{(rid,\ d,\ side)\ &|\ (rid,\ d) \in RMeas(R),\\
&side \in \{up,\ down,\ none\}\}.
  \end{split}
\]
The $side$ value indicates for $dual$ routes if a position can be reached
from the $up$ or the $down$ side of the route. For routes with $kind = simple$ the
$side$ value is always $none$.

Let $N=\{N_1,\ldots,N_k\}$ be a set of networks. A single network position in a
network $N_i$ is represented by the data type \dt{gpoint}. The carrier set of
\dt{gpoint} is defined as
\[\{(i,\ gp)\ |\ 1 \leq i \leq k \wedge gp \in RLoc(R) \cup \{ \perp \}\},\]
where $\perp$ again represents a undefined value.

A $route\ interval$ in $N_i$ is a pair of network positions on the same route. The $route\ interval$ is represented by a quadruple $(rid,\ d_1,\ d_2,\ side)$, with
$(rid,\ d_1,\ side),\ (rid,\ d_2,\ side) \in Loc(N)$
and $d_1 \leq d_2$. Semantically
a $route interval$ represents all route locations $(rid,\ d,\ side)$ with
$d_1 \leq d \leq d_2$. A finite set of disjoint $route\ intervals$ of a network
$N$ is called region of $N$. The set of all possible regions in a network $N$ is
denoted as $Reg(N)$.

A region within an network $N_i$ is represented by the data type \dt{gline}.
The carrier set of \dt{gline} is defined as
\[\{(i,\ gl)\ |\ 1 \leq i \leq k \wedge gl \in Reg(N_i) \}\].
The set of $route intervals$ defining a network region may be empty.

The type systesm of \cite{RepresentingMovingObjectsGueting} is extended by
\cite{NetworkGueting} to contain a new kind GRAPH consisting of the data types
\dt{gpoint} and \dt{gline}. The type constructors \dt{moving} is also extended
to be feasible for data types of kind GRAPH. Therefore the data type
\dt{moving}(\dt{gpoint}) (\dt{mgpoint} for short) is defined similar to the
\dt{mpoint} explained in detail in Section \ref{sec:bmodbdatamod}. The units of
a \dt{mgpoint} consist of \dt{ugpoint} values and single positions can be given
by \dt{igpoint} values.

The paper provides numerous operations on the network and the network data types.
We give reviews of the operations that were used in our experiments later in this
paper (see sections \ref{sec:implNDM} - \ref{sec:Translation}).

\cite{NetworkGueting} proposes also implementational issues for NET in
\secondo{}. The implementation of the data type \dt{network} consists of three
relations called $routes$, $junctions$, and $sections$, and a persistent adjacency
list data structure supporting trip and path computations.

>>>>

\documentclass[a4paper]{article}
%\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{float}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{todonotes}
\usepackage{rotating}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data
models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two
representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data
models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis
,HERMESMDCPelekis}
for DMFS and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,
NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

Why do we spend time on NCDM, if everything can be represented by DMFS?
Now, it is natural to give positions related to the street
network instead of coordinate pairs in the xy plane. NCDM are expected to use
less storage space,
because geographical information's about street curves are stored only once in
the network, whereas in DMFS each street curve is stored in each moving point
object using this street. NCDM can support query processing with specialized indexes
using their knowledge of the underlying network. It is much easier to
formulate queries about the relationships between moving objects and the network
in the NCDM. And not at last, the results of our experiments show that
our network constrained data model outperforms our data model of free movement in
two dimensional space by orders of magnitude. The network constrained data model
uses less than 60\% of the storage space and less than 50\% of the total query
run time of the data model of free movement in space, which we used in our
experiments.
We think that these results show that it is useful to develop specialized data
models for specialized data structures like NCDM for
network constrained moving objects to save storage space and reduce query run times.

@@

\cite{MOLRaQSu} proposes an incomplete abstract DMFS. Basic idea is that
spatio-temporal data types can be modeled by linear constraints and queries
can be formulated using formulas from differential geometry.

\cite{RepresentingMovingObjectsGueting} proposes an abstract DMFS including the
idea of an time sliced representation for moving objects. This basic idea of
time sliced representation is used by the DMFS \cite{STAUPelekis} and
\cite{DataModelDataStructureGueting}. The last one is used in our experiments
and therefore reviewed in more detail in Section \ref{sec:bmodbdatamod}.
The main difference between the both data models is that
\cite{DataModelDataStructureGueting} supports only linear interpolation of movement,
whereas \cite{STAUPelekis} also supports arc interpolation of movement.
\cite{STAUPelekis} uses only one spatial object
containing all spatial geometries for the representation of spatial objects and
one moving object for the representation of the different moving object data types,
whereas \cite{DataModelDataStructureGueting} uses different spatial and moving
objects for the representation of the different spatial and moving data types.
According to this, \cite{STAUPelekis} provides only a single operator that
distinguishes
between the different topological relationships via a parameter, whereas
\cite{DataModelDataStructureGueting} uses different operations to estimate
topological relationships. Overall, \cite{STAUPelekis} offers a more flexible
object oriented design than \cite{DataModelDataStructureGueting}.

@@

The most NCDM use edge based graph representations for the representation of the
underlying network data only a small number of NCDM use route oriented data
models or combine route and edge based data models.

\cite{NetworkVazirgiannis} proposes an edge based NCDM. The edges and their
attributes
are stored in a relation representing the network as an undirected graph. Moving
objects are assumed to drive always on the path with the lowest cost,
in terms of distance or travel-time. They are defined by the source point, the
target point, and the starting time instant of the trip. The trajectory is
computed by this assumption using the length and speed attributes of the graph
edges within the shortest, respectively, fastest path computation.
The advantage of this definition is the reduced storage space for the representation
of moving point objects. The drawback is the high computational effort for query
evaluation on moving objects.

@@

Another two-layer network representation is proposed by
\cite{DynamicTransportNetworkDing}.
The authors of \cite{DynamicTransportNetworkDing} combine the advantages of the
dynamic edge-based \cite{DynNetworkModEdgeGueting} and the dynamic route-based
\cite{DynNetworkModRouteGueting} NCDM approaches. The route-based environment
reduces the update intervals and used storage space for the database representation
of moving point objects, whereas the edge-based environment supports
a more detailed view on the traffic conditions of the different edges belonging
to the same route. Moving objects are represented by a set of pairs. The pairs
consist of an motion vector and an Boolean flag. The Boolean flag tells if the
motion vector contains current or historical information. Each motion vector
consists, in parts similar to \cite{NetworkJensen}, of a time stamp, a network
position, and a speed vector. Similar to \cite{NetworkGueting} the network
positions are given related to routes and junctions and not on edges. Different
from all
other NCDM in this section \cite{DynamicTransportNetworkDing}
uses time depending dynamic attributes in the representation of the network parts.
Therefore, changes in the network environment can be handled without loss of
information in this NCDM.

To the best of our knowledge only a few of the proposed data models have been
implemented into database management systems: \cite{STAUPelekis} is implemented
as data cartridge \cite{HERMESMDCPelekis, HERMESPelekis} for the commercial
Oracle\textregistered{} object-relational DBMS \cite{oracle};
\cite{DynamicTransportNetworkDing} is implemented as extension of the open source
database project PostgreSQL \cite{PostgreSQL}; \cite{DataModelDataStructureGueting}
and \cite{NetworkGueting} are implemented in the freely available extensible
\secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}.

Although \cite{STAUPelekis} for the DMFS and \cite{DynamicTransportNetworkDing} for
the NCDM provide greater flexibility we decided to use
\cite{DataModelDataStructureGueting} and \cite{NetworkGueting} in our
experiments, because both data models are available in the same DBMS, which is
also the DBMS in which the \bmodb{} \cite{BerlinMODVLDBDuentgen} has been
developed.

@@

Therefore, several data generators have been developed. Some of them generate
only unconstrained moving point objects like \cite{OportoSaglio}, or support
only short-term observations like \cite{BrinkhoffsDataGenerator}, or require
additional software like \cite{STACTS}.

In the sequel we give short reviews of the \secondo{} DBMS
(Section \ref{sec:secondo}), the both data models we used in our experiments
(Section \ref{sec:bmodbdatamod} and Section \ref{sec:netdatamod}), and the \bmodb{}
(Section \ref{sec:bmodb}).
\subsection{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use these plugins to extend
their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including SPACE (Section \ref{sec:bmodbdatamod}) and NET
(Section \ref{sec:netdatamod}). Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version 3.0.
\subsection{Data Model of \bmodb{} (SPACE)}
\label{sec:bmodbdatamod}
\cite{STDTModelingQueryingMOinDErwig} presents the basic idea of the DMFS that
is used by the \bmodb{}. The abstract data model for SPACE was pubished in
\cite{RepresentingMovingObjectsGueting} and the discrete data model in
\cite{DataModelDataStructureGueting}. Abstract data models are useful as conceptual
models, but they cannot be implemented, because computers can only use finite sets.
In the sequel we will mainly focus on the discrete data model.

@@

The carrier sets for all data types in the discrete data model contain
$\{\perp\}$. $\perp$ represents an undefined value.

The carrier sets for the BASE data types in the discrete data model are defined
by the corresponding programming language data types \dt{int}, \dt{real}, \dt{bool},
and \dt{string}. The carrier set of a value of the data type \dt{point} is
\dt{real} $\times$ \dt{real}. The two \dt{real} values represent the
coordinate pair of the position of the \dt{point} value in the xy plane.

The data type \dt{points} consists of a disjoint set of \dt{point} values.

The carrier set for the data type \dt{line}
consists of a finite set of disjoint line segments representing the linear
approximation of the line curve in the two dimensional plane. Semanticaly an
\dt{line} value is the union of the points of all its line segments.

A region
is the union of all points covered by the region. The carrier set \dt{region}
is defined to be a finite set of line segments building a polygon representing
the linear approximation of the outer and, if the region contains wholes, inner
borders of the region. The borders are defined to belong to the region.

@@

The type constructor \dt{range} converts BASE and TIME data types $\alpha$ into
a type whoes values are finite sets of intervals over $\alpha$. Range types are
used to represent collections of time intervals, or the values taken by a moving
real. Intervals are represented by their start and end point and two flags
indicating if the start, respectively, end point is part of the interval or not.

The other important type constructor of the abstract data model is \dt{moving}.
In the abstract data model \dt{moving} maps each data type $\alpha$ from BASE
and
SPATIAL into an time dependend spatio-temporal moving data type
\dt{moving}($\alpha$)
(\dt{m$\alpha$} for short) of kind TEMPORAL. The discrete data model introduces some
additional type constructors to implement the moving type constructor of the
abstract data model. A detailed description of the type constructors is skiped
due to place limitations. We explain the realisation of \dt{moving} at the example
of a \dt{moving}(\dt{point}) (short \dt{mpoint}) object. An \dt{mpoint} value
may represent a car, which changes its position in the plain within time.

An \dt{mpoint} consists of a set of so called \dt{unit}(\dt{point}) values
(\dt{upoint} for short).
Each \dt{upoint} consists of a time interval and two \dt{point} values. The first
\dt{point} value represents the position of the \dt{upoint} at the start of the
time interval and the second \dt{point} value represents the position of the
\dt{upoint} at the end of the time interval. It is assumed that the object,
represented by the \dt{upoint}, moves on the straight line between these two
points with constant
speed within the given time interval. The velocity of the object is given by the
ratio from the distance of the two points and the length of the time interval of
the \dt{upoint}.

@@

This spatio-temporal data model of \dt{moving} allows us to compute the position
of an \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assuming the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by an \dt{intime}(\dt{point})
(short form \dt{ipoint}). An \dt{ipoint} consists of an time instant and an
\dt{point}
value and represents the position of the \dt{mpoint} value at the given time
instant.

Other data types of \secondo{} which are used in the \bmodb{} are \dt{mbool},
\dt{mreal}, and \dt{periods}. A \dt{mbool} value consists of a set of \dt{ubool}
values. Each \dt{ubool} value is is constant \true{} or \false{} for the given
time interval. A \dt{mreal} value consists of a set of \dt{ureal} values. Each
\dt{ureal} value is defined by a function of time representing the \dt{real}
value at each time instant. A \dt{periods} value is a set of disjoint an not
connected time intervals.
\subsection{Network Data Model(NET)}
\label{sec:netdatamod}
The central idea of the network constrained data model NET presented in
\cite{NetworkGueting} is that
every movement is constrained by a given network and every position can be described
relative to this network. Contrary to the most other NCDM NET models the network
in terms of routes, corresponding to roads or highways in real life. Positions
are given by a route identifier and the distance from the start of the route.
This is a more natural representation of network positions as the directed graph
representation of networks, where junctions are vertexes and the pieces between
junctions are represented by edges, which is used in the most NCDM. We have names
for roads not for junctions or pieces between junctions.

The routes based network representation has the advantage that the
representation of moving objects that move over several sections of the same route
with constant speed becomes much smaller. We only have to store an new unit
if the moving object changes the route or the speed, not every time
it passes a junction like in the edge based NCDM.

In NET the data type \dt{network} is modeled by two main components. One is the
set  of routes (streets) and the other one the set of junctions (crossings). The
domain of routes is defined as
\[
\begin{split}
Route = \{(id,\ l,\ c,\ kind,\ start)\ &|\ id \in \dt{int},\ l \in \dt{real},
\ c \in \dt{line},\\
&\ kind \in \{simple,\ dual\},\\
&\ start \in \{smaller,\ larger\}\},
\end{split}
\]
where $id$ is a distinct route idenitfier, $l$ is the length of the route, $c$
is the route curve as \dt{line} value (see Section \ref{sec:bmodbdatamod}),
\textit{kind} indicates if the lanes of the route are separated, and \textit{start} indicates
how the route curve is embedded into space.

If $R$ is a set of distinct routes, the domain of junctions in $R$ is defined as
\[
  \begin{split}
    Junction(R)=\{(rm_1,\ rm_2,\ cc)\ &|\ rm_1,\ rm_2 \in RMeas(R),\\
&rm_1 = (r_1,\ d_1),\ rm_2 = (r_2,\ d_2),\\
&r_1 \neq r_2,\ cc \in \dt{int}\}.
  \end{split}
\]
Where the set of possible positions in $R$ $RMeas(R)$ is defined as
\[
  \begin{split}
    RMeas(R) = \{(rid, d)\ &|\ rid \in \dt{int},\ d \in \dt{real},\\
&\exists (rid,\ l,\ c,\ k,\ s) \in R \wedge 0 \leq d \leq l\},
  \end{split}
\]
where $d$ is the distance from the start of the route and the connectivity code
$cc$ encodes which lanes of the routes are connected
by the junction\footnote{See \cite{NetworkGueting} for a detailed
explanation of the different connectivity code values.}.

A network $N$ is a pair $(R,\ J)$, where $R$ is a finite set of distinct routes
and $J$ is a finite set of junctions in $R$. The carrier set for network
positions $Loc(N)$ is equal to the set of route locations $RLoc(R)$ and defined
as
\[
  \begin{split}
    RLoc(R)=\{(rid,\ d,\ side)\ &|\ (rid,\ d) \in RMeas(R),\\
&side \in \{up,\ down,\ none\}\}.
  \end{split}
\]
The \textit{side} value indicates for \textit{dual} routes if a position can be reached
from the \textit{up} or the \textit{down} side of the route. For routes of \textit{kind simple}
the \textit{side} value is always \textit{none}.

Let $N=\{N_1,\ldots,N_k\}$ be a set of networks. A single network position in a
network $N_i$ is represented by the data type \dt{gpoint}. The carrier set of
\dt{gpoint} is defined as
\[\{(i,\ gp)\ |\ 1 \leq i \leq k \wedge gp \in RLoc(R) \cup \{ \perp \}\},\]
where $\perp$ again represents an undefined value.

A \textit{route interval} in $N_i$ is a pair of network positions on the same
route. 
The \textit{route interval} is represented by a quadruple $(rid,\ d_1,\ d_2,\
side)$, with
$(rid,\ d_1,\ side),\ (rid,\ d_2,\ side) \in Loc(N)$
and $d_1 \leq d_2$. Semantically
a \textit{route interval} represents all route locations $(rid,\ d,\ side)$ with
$d_1 \leq d \leq d_2$. A finite set of disjoint \textit{route intervals} of a
network
$N_i$ is called region of $N_i$. The set of all possible regions in a network
$N_i$ is
denoted as $Reg(N_i)$.

A region within an network $N_i$ is represented by the data type \dt{gline}.
The carrier set of \dt{gline} is defined as
\[\{(i,\ gl)\ |\ 1 \leq i \leq k \wedge gl \in Reg(N_i) \} \cup \{\}\].
The set of \textit{route intervals} defining a network region may be empty.

The type systesm of \cite{RepresentingMovingObjectsGueting} is extended by
\cite{NetworkGueting} to contain a new kind GRAPH consisting of the data types
\dt{gpoint} and \dt{gline}. The type constructors \dt{moving} is also extended
to be feasible for data types of kind GRAPH. Therefore the data type
\dt{moving}(\dt{gpoint}) (\dt{mgpoint} for short) is defined similar to the
\dt{mpoint} explained in detail in Section \ref{sec:bmodbdatamod}. The units of
a \dt{mgpoint} consist of \dt{ugpoint} values and single positions of a
\dt{mgpoint} are be given by \dt{igpoint} values.

The paper provides numerous operations on the network and the network data
types.
We give reviews of the operations that were use in our experiments later in
Section \ref{sec:implNDM} and Section \ref{sec:Translation}.
